\textbf{The Heart can Lie: A Preliminary Investigation of the Role of Interoception and Theory of Mind in Deception}

### 2023-10-02 paper/manuscript.tex
update

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,mask,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Deception, Interoception, Theory of Mind, Poygraph, Lying Ability\newline\indent Word count: 3720}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage[labelfont=bf, font={color=gray,small}]{caption}
\usepackage{float}
\usepackage[document]{ragged2e}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The Heart can Lie: The Role of Interoception and Theory of Mind in Deception},
  pdflang={en-EN},
  pdfkeywords={Deception, Interoception, Theory of Mind, Poygraph, Lying Ability},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{The Heart can Lie: The Role of Interoception and Theory of Mind in Deception}}
\author{Dominique Makowski\textsuperscript{1, 2}, Zen J. Lau\textsuperscript{2}, Tam Pham\textsuperscript{2}, An Shu Te\textsuperscript{2}, Stephanie Kirk\textsuperscript{2}, Claudia Liauw\textsuperscript{2}, \& S.H. Annabel Chen\textsuperscript{2, 3, 4, 5}}
\date{}

@@

The authors made the following contributions. Dominique Makowski: Conceptualization, Data curation, Formal Analysis, Funding acquisition, Investigation, Methodology, Project administration, Resources, Software, Supervision, Validation, Visualization, Writing -- original draft; An Shu Te: Project administration, Resources, Investigation, Writing -- original draft; Stephanie Kirk: Project administration, Resources, Writing -- original draft; Claudia Liauw: Data curation, Formal Analysis; S.H. Annabel Chen: Project administration, Supervision, Writing -- review \& editing.

@@

\abstract{%
While a large part of the deception literature focuses on lying detection, the factors contributing to one's ability to lie remain unclear. The present study examined the contribution of Theory of Mind (ToM) and interoception on our ability to lie using a directed lie paradigm with two conditions (``Interrogation'' and ``Polygraph''), designed to enhance each of the two mechanisms. Using a mixed-design Bayesian analysis, we show that various facets of interoceptive abilities are positively related to the self-rated confidence in one's own lies, especially when under the belief that bodily signals are being monitored (i.e., in the ``Polygraph'' condition). Beyond providing evidence for the role of the body in lying and raising interesting questions for deception science, these results carry practical implications for criminology and lie detection protocols.
}

@@

Lying - the intentional attempt at instilling a false belief in others (Sip et al., 2012) - is a prevalent phenomenon carrying potentially important consequences. Interestingly, evidence suggests that the successful detection of a lying attempt depends more on the ability of the liar, than on the performance of the lie detector (Bond Jr \& DePaulo, 2008; Levine et al., 2011; Verigin et al., 2019). However, with most of the deception literature focused on deception detection (Masip, 2017; Sternglanz et al., 2019; Viji et al., 2022), the factors contributing to one's ability to lie remain unclear.

@@

Besides paying attention to the person we lie to, gauging whether they believe us, some attention is also directed inwards: monitoring our own body and its reactions (e.g., cardiac activity and its related changes such as blushing), which could be used as cues to infer our real intent. This begs the question of the potential role of interoceptive abilities in deception ability. Broadly defined as one's sensitivity to their own internal signals and bodily states (Chen et al., 2021; Murphy et al., 2019; Weiss et al., 2014), Garfinkel et al. (2015)'s conceptualizes interoception as a three-dimensional construct comprising three distinct facets, namely, interoceptive accuracy - the objective ability to monitor internal bodily signals; interoceptive sensibility - the subjective confidence in one's interoceptive accuracy; and interoceptive awareness - one's ability to identify and appropriately respond to their perceived interoceptive states. Interoception has increasingly been tied to subjective perceptual experiences (Connell et al., 2018; Seth et al., 2012), as well as individual differences in executive functions, emotional processing, and decision-making (Barrett \& Simmons, 2015; Murphy et al., 2019; Petzschner et al., 2021). Although few studies exist that investigate the relationship between interoception and deceptive ability \emph{per se}, previous decision-making studies have demonstrated a negative correlation between interoceptive awareness, a metacognitive dimension of interoception, and one's likelihood to make risky decisions (Dunn et al., 2010; Furman et al., 2013). In fact, this is consistent with the somatic marker hypothesis, which posits an association between interoception and a propensity towards making rational decisions (Damasio, 1996).

In contrast, however, some studies have instead found heightened interoceptive attention (one's self-focus towards internal bodily signals), to predict apathetic, immoral behaviour, such as cheating (Ditto et al., 2006; Lenggenhager et al., 2013; Williams et al., 2016). Extending these findings to social cognition, Vabba et al. (2022) further reports individuals with lower interoception told significantly less egoistic lies when the social reputational stakes were high, whereas individuals with higher interoception did not exhibit a significant difference in the number of lies told. Given the scarce research on interoception and deception, more studies are herein needed to clarify these mixed findings.

The aim of the present study was to explore the contribution of ToM and interoception abilities on individuals' deception skills, as indicated by their lying confidence, physiological arousal and response time. To this end, we designed a directed-lying paradigm with 2 conditions differing in the nature of their feedback cues. The \emph{Interrogation} condition was designed to emphasize (and preferentially mobilize) ToM-related mechanisms, whereas the \emph{Polygraph} condition was designed to emphasize interoceptive mechanisms. We expected ToM and interoception to positively predict lying ability (i.e., higher lie confidence, shorter response time and lower physiological arousal), in particular in the \emph{Polygraph} and the \emph{Interrogation} condition, respectively.

@@

30 university students from Singapore were recruited through posters, flyers, and online social media platforms.
Four participants were excluded as their data was not recorded due to technical issues. The final sample consists 26 participants (Mean age = 20.9, SD = 2.0, range:{[}18, 25{]}, Sex: 65.4\% women, 34.6\% men). The heart rate of one participant and response time of one participant were excluded from further analysis due to extreme outlying values.

@@

\hypertarget{theory-of-mind}{%
\subsubsection{Theory of Mind}\label{theory-of-mind}}

Two measures of ToM and its related constructs were administered. The Yoni Task (Shamay-Tsoory \& Aharon-Peretz, 2007) is a behavioral task in which participants are presented with the face of a character named ``Yoni'', surrounded by 4 colored pictures of objects or faces. In total, each participant completed 101 trials - 49 trials assessing their affective TOM abilities, 37 trials assessing their cognitive TOM abilities and 15 control trials. During each trial, participants were shown a question and asked to make responses based on specific cues such as directions of Yoni's eye gaze, facial expressions etc., In the control trials, participants made judgements based on Yoni's physical context (physical TOM).

The Basic Empathy Scale (BES, Jolliffe \& Farrington, 2006), a 20-item questionnaire measuring two dimensions of empathy (cognitive and affective) using a 5-point Likert scale was administered. Although ToM and empathy are regarded as distinct psychological constructs, previous research findings point to them being closely related (Gallant et al., 2020; Sebastian et al., 2012).

@@

The MAIA-2 (Mehling et al., 2012), a 32-item questionnaire which measures 8 dimensions of interoception (Noticing, Not-Distracting, Not-Worrying, Attention, Regulation, Emotional Awareness, Self-Regulation, Body Listening, and Trust) using 5-point Likert scales, was also administered.

@@

Using PsychoPy (Peirce et al., 2019), we implemented a directed-lying task in which participants were instructed to answer 80 questions by either lying or telling the truth (depending on whether they see ``lie'' or ``truth'' written on the screen). Importantly, they were told that for half of the trials, they would have to convince another participant that would be observing them from a separate room via a webcam connection (\emph{Interrogation} condition). For the remaining trials, participants were tasked to convince a ``lie detection machine'' that would be assessing their behaviour through their physiological signals (\emph{Polygraph} condition). The two conditions were presented in a counter-balanced order, and each comprised of 40 trials (20 truth; 20 lies).

The sequence of each trial was the same for both conditions. Participants were instructed to respond to a question shown on a computer screen, phrased either directly (e.g., ``What is your favourite sport?''), or indirectly (e.g., ``Is your favourite sport Hockey?''), by verbally lying or telling the truth (for half of the trials in each condition, i.e., n = 20). Following a short interval (0.7 - 1.5s) for response preparation, during which ``Connecting\ldots{}'' was shown on the screen, a social or bio-feedback cue (for \emph{Interrogation} and \emph{Polygraph} conditions respectively) was displayed for a maximum of 10s. Participants then had to provide their answers verbally, and pressed the space key to signify the end of their response. After another short interval (1.5-2.5s), during which ``Disconnecting'' was written on the screen, participants were asked to rate their confidence in how convincing (i.e., likely to get judged as truthful) they perceived their response to be on a visual analog scale.

@@

In the \emph{Polygraph} condition, participants had to provide their answer while receiving biofeedback in the form of physiological signals (ECG, EDA and Respiration) being recorded. Participants were informed that they were being shown live feedback of their own physiological activity. In actuality, we displayed pre-recorded video clips of a confederate's physiological signals. All participants reported believing that it was indeed their own signals being shown during the debriefing.

Finally, on top of the deception task, we also measured participants' self-reported tendency to lie in their everyday life using the \emph{Lie Scale} (Makowski, Pham, Lau, Raine, et al., 2021), a 16-item questionnaire that assesses 4 dispositional lying dimensions (Ability, Negativity, Contextuality and Frequency).

@@

This study is comprised of 2 sessions. During session 1, participants answered a brief demographic survey before the questionnaires, which were presented in a randomized order. During session 2, cognitive-behavioural tasks were administered to participants while their physiological signals (including cardiac activity (ECG), respiration (RSP) and electrodermal activity (EDA)) were being recorded. The physiological recording devices were set up as follows: ECG was recorded with 3 electrodes placed according to a modified Lead II configuration (Takuma et al., 1995), and respiration was measured using a respiration belt. All signals were recorded at 1000Hz via the BioPac MP160 system (BioPac Systems Inc., USA). For all participants, session 2 began with the deception task, followed by the YONI task and the HCT, with the latter two presented in a randomized order. Three outcome variables were recorded for each trial of the deception task, namely the participants' confidence ratings that their answers (lies \emph{vs.} truths) were convincing, the response time (RT) between the question onset and the participant's key press (indicating the end of their verbal answer), and the heart rate change associated with the response (within a window of 3.5 s).

@@

To maximize the signal-to-noise ratio, we performed a feature reduction on our two groups of predictor variables (namely, ToM and interoception) using factor analysis over PCA, as the goal was to extract meaningful and consistent factors, rather than merely maximizing the variance explained. Then, we modelled the relationship between these inter-individual composite scores (note that the analysis for all individual variables is nonetheless included in the analysis report) and the 3 outcome variables in interaction with the condition (polygraph \emph{vs.} interrogation). Finally, we investigated the relationship between the deception scale traits, and the ToM and interoception scores using Bayesian correlations. As all the analyses and data has been made available, we will in the manuscript focus on significant, i.e., - in this context - statistically reliable and in our opinion theoretically relevant.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,mask,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Deception; Interoception; Theory of Mind; Poygraph; Lying Ability\newline\indent Word count: 5004}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage[labelfont=bf, font={color=gray,small}]{caption}
\usepackage{float}
\usepackage[document]{ragged2e}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The Heart can Lie: A Preliminary Investigation of the Role of Interoception and Theory of Mind in Deception},
  pdflang={en-EN},
  pdfkeywords={Deception; Interoception; Theory of Mind; Poygraph; Lying Ability},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{The Heart can Lie: A Preliminary Investigation of the Role of Interoception and Theory of Mind in Deception}}
\author{Dominique Makowski\textsuperscript{1, 2}, Zen J. Lau\textsuperscript{2}, Tam Pham\textsuperscript{2}, An Shu Te\textsuperscript{2}, Stephanie Kirk\textsuperscript{2}, Claudia Liauw\textsuperscript{2}, \& S.H. Annabel Chen\textsuperscript{2, 3, 4, 5}}
\date{}

@@

The authors made the following contributions. Dominique Makowski: Conceptualization, Data curation, Formal Analysis, Funding acquisition, Investigation, Methodology, Project administration, Resources, Software, Supervision, Validation, Visualization, Writing -- original draft; Zen J. Lau: Data curation, Project administration, Resources, Investigation; Tam Pham: Data curation, Project administration, Resources, Investigation; An Shu Te: Project administration, Resources, Investigation, Writing -- original draft; Stephanie Kirk: Project administration, Resources, Writing -- original draft; Claudia Liauw: Data curation, Formal Analysis; S.H. Annabel Chen: Project administration, Supervision, Writing -- review \& editing.

@@

\abstract{%
While a large part of the deception literature focuses on lying detection, the factors contributing to one's ability to lie remain unclear. The present study examined the contribution of Theory of Mind (ToM) and interoception on our ability to lie using a directed lie paradigm with two conditions (``Interrogation'' and ``Polygraph''), designed to enhance each of the two mechanisms. Given the relatively small sample size (n = 26 x 40 trials), special steps were taken to avoid false positives. Our results suggest that various facets of interoceptive abilities are positively related to the self-rated confidence in one's own lies, especially when under the belief that bodily signals are being monitored (i.e., in the ``Polygraph'' condition). Beyond providing evidence for the role of the body in lying and raising interesting questions for deception science, these results carry practical implications for criminology and lie detection protocols.
}

@@

Lying - the intentional attempt at instilling a false belief in others (Sip et al., 2012) - is a prevalent phenomenon carrying potentially important consequences. Interestingly, evidence suggests that the successful detection of a lying attempt depends more on the ability of the liar, than on the performance of the lie detector (Bond Jr \& DePaulo, 2008; T. R. Levine et al., 2011; Verigin et al., 2019). However, with most of the deception literature focused on deception detection (Masip, 2017; Sternglanz et al., 2019; Viji et al., 2022), the factors contributing to one's ability to lie remain unclear. Nevertheless, some findings suggest a relationship between the propensity to tell lies, and traits that characterize the socially malevolent profile known as the Dark Triad (Paulhus \& Williams, 2002), such as narcissism (Zvi \& Elaad, 2018) and psychopathy (Rassin et al., 2023). While often conceptualized to be immoral and unconscionable, lying is ubiquitous in everyday life, and being able to lie skillfully can sometimes facilitate interpersonal relationships, helping us avoid conflict or causing emotional harm to others (E. E. Levine \& Lupoli, 2022). In fact, recent research shows that certain forms of deception, such as prosocial lies (i.e., false statements told to benefit others; E. E. Levine and Lupoli (2022)), can increase trust (E. E. Levine \& Schweitzer, 2015). Moreover, individuals who told altruistic lies were perceived as more benevolent than those who were honest (E. E. Levine \& Schweitzer, 2014).

@@

Besides paying attention to the person we lie to, gauging whether they believe us, some attention is also directed inwards: monitoring our own body and its reactions (e.g., cardiac activity and its related changes such as blushing), which could be used as cues to infer our real intent. This begs the question of the potential role of interoceptive abilities in deception ability. Broadly defined as one's sensitivity to their own internal signals and bodily states (Chen et al., 2021; Murphy et al., 2019; Weiss et al., 2014), Garfinkel et al. (2015) conceptualize interoception as a three-dimensional construct comprising three distinct facets, namely, interoceptive accuracy - the objective ability to monitor internal bodily signals; interoceptive sensibility - the subjective confidence in one's interoceptive accuracy; and interoceptive awareness - the metacognitive ability to correctly evaluate one's interoceptive ability. Interoception has increasingly been tied to subjective perceptual experiences (Connell et al., 2018; Seth et al., 2012), as well as individual differences in executive functions, emotional processing, and decision-making (Barrett \& Simmons, 2015; Murphy et al., 2019; Petzschner et al., 2021).

Although few studies exist that investigate the relationship between interoception and deceptive ability \emph{per se}, previous decision-making studies have demonstrated a negative correlation between interoceptive awareness, a metacognitive dimension of interoception, and one's likelihood to make risky decisions (Dunn et al., 2010; Furman et al., 2013). This is in line with the somatic marker hypothesis, which posits that an accurate evaluation of one's bodily signals facilitates the use of such interoceptive feedback to guide rational decision making (Damasio, 1996). Indeed, Sugawara et al. (2020) further reports that individuals who received interoceptive training were more likely to show higher interoceptive accuracy and make reasoned decisions. Given that deciding to lie generally involves a consideration of the potential costs of getting caught, and hence could also be perceived as risky behavior (Kireev et al., 2013), interoception could be construed to be negatively related to lying ability. However, some studies have instead found heightened interoceptive attention (one's self-focus towards internal bodily signals), to predict immoral behaviour, such as cheating (Ditto et al., 2006; Lenggenhager et al., 2013; Williams et al., 2016). Extending these findings to social cognition, Vabba et al. (2022) further reports individuals with lower interoception told significantly less egoistic lies when the social reputational stakes were high, whereas individuals with higher interoception did not exhibit a significant difference in the number of lies told. Given the scarce research on interoception and deception, more studies are herein needed to clarify these mixed findings.

The aim of the present study was to explore the contribution of ToM and interoception abilities on individuals' deception skills, as indicated by their lying confidence, physiological arousal and response time. To this end, we designed a directed-lying paradigm with 2 conditions differing in the nature of their feedback cues. The \emph{Interrogation} condition was designed to emphasize (and preferentially mobilize) ToM-related mechanisms, whereas the \emph{Polygraph} condition was designed to emphasize interoceptive mechanisms. In particular, we expected lying ability (i.e., higher lie confidence, shorter response time and lower physiological arousal), to be positively predicted by individuals' interoceptive abilities in the \emph{Polygraph} condition, and by ToM skills in the \emph{Interrogation} condition. Consistent with the cognitive load approach outlined in several theories of deception (such as the Four-Factor Theory (Riggio et al., 1987) and Activation-Decision-Construction Model (Walczyk et al., 2014)), as well as previous findings which suggest response time as a reliable cue to deception (Gonzalez-Billandon et al., 2019; Walczyk et al., 2009), we regarded shorter response times as a proxy of better lying ability.

@@

30 university students from Singapore were recruited through posters, flyers, and online social media platforms, and rewarded with study credits for their time.
Four participants were excluded as their data was not recorded due to technical issues. The final sample consists of 26 participants (Mean age = 20.9, SD = 2.0, range:{[}18, 25{]}, Sex: 65.4\% women, 34.6\% men). The heart rate of one participant and response time of one participant were excluded from further analysis due to extreme outlying values. To maximize statistical power, the problematic data from these 2 participants were only excluded from analyses involving those measures; all other data were retained for analyses.

@@

\hypertarget{theory-of-mind-tom}{%
\subsubsection{Theory of Mind (ToM)}\label{theory-of-mind-tom}}

Two measures of ToM and its related constructs were administered. The Yoni Task (Shamay-Tsoory \& Aharon-Peretz, 2007) is a behavioral task which assesses first and second-order ToM abilities in both cognitive and affective domains. Participants were presented with the face of a character named ``Yoni'', surrounded by 4 colored pictures of objects or faces, one in each corner of the screen. In total, each participant completed 101 trials - 49 trials assessing their affective ToM abilities, 37 trials assessing their cognitive ToM abilities and 15 control trials. During each trial, participants were shown a question pertaining the item Yoni is referring to, and asked to make responses based on specific corresponding cues such as the directions of Yoni's eye gaze, facial expressions etc., In the control trials, participants made judgements based on Yoni's physical context (physical ToM). More specifically, in first-order trials, participants were instructed to make inferences about Yoni's mental state with regards to the objects surrounding it (e.g., ``Yoni is thinking of\ldots{}''). In more complex second-order trials, participants had to correctly infer the interaction between Yoni and others' mental states (e.g., Yoni is thinking of the fruit that \ldots{} wants'').

The Basic Empathy Scale (BES, Jolliffe \& Farrington, 2006), a 20-item self-report questionnaire measuring two dimensions of empathy, namely Cognitive (\(\alpha = 0.83\)) and Affective (\(\alpha = 0.82\)) using a 5-point Likert scale was administered. Although ToM and empathy are regarded as distinct psychological constructs, previous research findings point to them being closely related (Gallant et al., 2020; Sebastian et al., 2012). Specifically, empathy is often thought to be an integral component in the affective dimension of ToM (i.e., the ability to infer what someone else is feeling) (Shamay-Tsoory et al., 2010).

@@

Given its multidimensional nature, the MAIA-2 (Mehling et al., 2012), a 37-item questionnaire which measures 8 distinct facets of interoception including Noticing (e.g., \emph{I notice when I am uncomfortable in my body}; \(\alpha = 0.70\)), Not-Distracting (e.g., \emph{I try to ignore pain}; \(\alpha = 0.87\)), Not-Worrying (e.g., \emph{I can stay calm and not worry when I have feelings of discomfort or pain}; \(\alpha = 0.68\)), Attention Regulation (e.g., \emph{I can refocus my attention from thinking to sensing my body}; \(\alpha = 0.85\)), Emotional Awareness (e.g., \emph{I notice how my body changes when I am angry}; \(\alpha = 0.75\)), Self-Regulation (e.g., \emph{I can use my breath to reduce tension}; \(\alpha = 0.62\)), Body Listening (e.g., \emph{I listen to information from my body about my emotional state}; \(\alpha = 0.88\)), and Trust (e.g., \emph{I trust my body sensations}; \(\alpha = 0.89\)), using 5-point Likert scales, was also administered.

@@

Using PsychoPy (Peirce et al., 2019), we implemented a directed-lying task in which participants were instructed to briefly answer 80 questions (taken from the Autobiographical Memory Questionnaire - AMQ, Rubin et al., 2003) pertaining to their personal preferences and subjective experiences, by either lying or telling the truth (depending on whether they see ``lie'' or ``truth'' written on the screen). Their goal was to make convincing answers, so that truths would be judged as truths by the receiver, and lies as lies. The nature of the receiver was different depending on the condition: participants were told that for half of the trials, they would have to convince another participant that would be observing them from a separate room (COVID regulations were used as a justification) via a webcam connection (\emph{Interrogation} condition). For the remaining trials, participants were tasked to convince a ``lie detection machine'' that would be assessing their behaviour through their physiological signals (\emph{Polygraph} condition). In reality, there was no real ``receiver'' and their answers were not judged externally (the study focused on their subjective ratings and reactions). The two conditions were presented in a counter-balanced order, and each comprised of 40 trials (20 truth; 20 lies).

The sequence of each trial was the same for both conditions. Participants were first instructed to respond to a question shown on a computer screen by verbally lying or telling the truth (for half of the trials in each condition, i.e., n = 20). In addition, as past studies have found associations between lying behaviour and type of question phrasing (Walczyk \& Cockrell, 2022), each question was phrased either directly (e.g., ``What is your favourite sport?''), or indirectly (e.g., ``Is your favourite sport Hockey?'') to reduce possible confounding effects. Following a short interval (0.7 - 1.5s) to allow time for response preparation, during which ``Connecting\ldots{}'' was shown on the screen, a social or bio-feedback cue (for \emph{Interrogation} and \emph{Polygraph} conditions respectively) was displayed for a maximum of 10s or until a response was given. Specifically, participants had to provide their answers verbally, and pressed the space key to signify the end of their response. After another short interval (1.5-2.5s), during which ``Disconnecting\ldots{}'' was presented on the screen, participants were asked to rate their confidence in how convincing (i.e., likely to get judged as truthful) they perceived their response to be on a visual analog scale.

@@

In the \emph{Polygraph} condition, participants had to provide their answer while receiving bio-feedback in the form of physiological signals (ECG, EDA and Respiration), of which they were informed was live feedback of their own physiological activity. In actuality, we displayed pre-recorded video clips of a confederate's physiological signals. All participants reported believing that their own signals were shown during the debriefing.

Three outcome variables were recorded for each trial of the deception task, namely the participants' confidence ratings that their answers (lies or truths) were convincing, the response time (RT) between the question onset and the participant's key press (indicating the end of their verbal answer), and the change in heart rate associated with the response (within a window of 3.5 s).

Finally, on top of the deception task, we also measured participants' self-reported tendency to lie in their everyday life using the \emph{Lie Scale} (Makowski, Pham, Lau, Raine, et al., 2021), a 16-item questionnaire that assesses 4 dispositional lying dimensions - Ability (\(\alpha = 0.92\)), Frequency (\(\alpha = 0.66\)), Negativity (\(\alpha = 0.66\)), and Contextuality (\(\alpha = 0.70\)).

@@

A within-subjects design was used in the present study, which is comprised of 2 sessions, to investigate the roles interoception and ToM play in lying ability. During session 1, participants answered a brief demographic survey as well as a questionnaire regarding their personal preferences and subjective experiences (Autobiographical Memory Questionnaire, AMQ), followed by a series of psychological scales (i.e., BES, MAIA and Lie scale), which were randomly displayed.

During session 2, performed about one week later, the 3 cognitive-behavioural tasks (i.e., the deception task, HCT and the Yoni task) were administered to participants while their physiological signals (including cardiac activity - ECG, respiration -RSP, and electrodermal activity - EDA) were being recorded. The physiological recording devices were set up as follows: ECG was recorded with 3 electrodes placed according to a modified Lead II configuration (Takuma et al., 1995), and respiration was measured using a respiration belt. All signals were recorded at 1000Hz via the BioPac MP160 system (BioPac Systems Inc., USA).

For all participants, session 2 began with the deception task, followed by the Yoni task and the HCT, with the latter two presented in a randomized order. In the directed-lying task, items of the AMQ were presented as stimuli, with participants' recorded responses (in session 1) used to establish the ground truth.

@@

To maximize the signal-to-noise ratio, we performed a feature reduction on our two groups of predictor variables (namely, ToM and interoception) using factor analysis over PCA, as the goal was to extract meaningful and consistent factors, rather than merely maximizing the variance explained. Then, we modelled the relationship between these inter-individual composite scores (note that the analysis for all individual variables is nonetheless included in the analysis report) and the 3 outcome variables in interaction with the condition (polygraph \emph{vs.} interrogation). Finally, we investigated the relationship between the deception scale traits, and the ToM and interoception scores using Bayesian correlations. As all the analyses and data have been made available, we will in the manuscript focus on significant, i.e., - in this context - statistically reliable and in our opinion theoretically relevant.
